---
title: "NBL1Manuscript_Figure2"
author: "Courtney Willey"
date: "2025-10-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Code Script for Nbl1 Manuscript: NBL1 Correlates with Renal Phenotypes in Mouse Models of Kidney Disease, but is Not Causal.

# Figure 2.  Testing Causality using an Nbl1 Knockout Mouse with XLAS.

Serum NBL1 levels were calcualted via ELISA at 6, 10, and 15 weeks of age. GFR was measured at 14-weeks and ACR was measured at 15-weeks. 

```{r Install}
# Install Packages
install.packages("ggplot2")
install.packages("readxl")
install.packages("dplyr")
install.packages("ggpubr")
install.packages("multcomp")
```

```{r Load}
# Load Packages
library(ggplot2)
library(readxl)
library(dplyr)
library(ggpubr)
library(multcomp)
```

# NBL1 Levels 

NBL1 levels at 6, 10, and 15 weeks of age. 

```{r NBL1}
# NBL1 Levels

# Load Data
NBL1Levels2206Combined <- read_excel("~/Library/CloudStorage/OneDrive-TheJacksonLaboratory/Korstanje Lab/22-06 NBL1 KO/22-06 Assays/NBL1Levels2206Combined.xlsx")

# Subset Data by Timepoint
NBL1Levels2206Combined$Timepoint <- factor(NBL1Levels2206Combined$Timepoint, levels=c("6w", "10w", "15w"))

# Calculate Summary Statistics
summary_stats <- NBL1Levels2206Combined %>%
  group_by(Timepoint, Genotype) %>%
  summarise(
    mean = mean(Nbl1, na.rm = TRUE),
    sem = sd(Nbl1, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

# Plot Results
ggplot(NBL1Levels2206Combined, aes(x = Genotype, y = Nbl1, color = Genotype)) +
  facet_grid(. ~ Timepoint) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.7) + # Dot plot
  geom_point(data = summary_stats, aes(y = mean), shape = 18, size = 4, color = "black") + # Mean
  geom_errorbar(data = summary_stats, aes(y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.2, color = "black") + # SEM
  scale_color_manual(values = c("orange", "purple")) +
  ylim(0, 125000) +
  ylab("NBL1 (pg/mL)") +
  ggtitle("NBL1 levels for each genotype across timepoints")

# Obtaining P-Values
# 6w
interaction6wData<-filter(NBL1Levels2206Combined, Timepoint=="6w")
interaction6W<-aov(Nbl1~Genotype,interaction6wData)
summary(interaction6W)
# p-value = 0.00163

#10w
interaction10wData<-filter(NBL1Levels2206Combined, Timepoint=="10w")
interaction10w<-aov(Nbl1~Genotype, interaction10wData)
summary(interaction10w)
# p-value = 0.0219

#15w
interaction15wData<-filter(NBL1Levels2206Combined, Timepoint=="15w")
interaction15w<-aov(Nbl1~Genotype, interaction15wData)
summary(interaction15w)
# p-value = 9.4e-05

# Function to calculate R² 
get_r2 <- function(model) {
  anova_table <- anova(model)
  ss_total <- sum(anova_table$"Sum Sq")
  ss_model <- sum(anova_table$"Sum Sq"[1])  # Only Genotype term
  r_squared <- ss_model / ss_total
  return(r_squared)
}

# Calculate R² for each timepoint
r2_6w <- get_r2(interaction6W)
r2_10w <- get_r2(interaction10w)
r2_15w <- get_r2(interaction15w)

# Print results
cat("R² for 6w:", r2_6w, "\n") # r2 = 0.4677964
cat("R² for 10w:", r2_10w, "\n") # r2 = 0.208233
cat("R² for 15w:", r2_15w, "\n") # r2 = 0.5241731
```

# Glomerular Filtration Rate (GFR) 

GFR Levels at 14 weeks of age.

```{r GFR}
# Glomerular Filtration Rate (GFR) @ 14 weeks

# Load Data
GFR_2206 <- read_excel("~/Library/CloudStorage/OneDrive-TheJacksonLaboratory/Korstanje Lab/22-06 NBL1 KO/22-06 Assays/GFR_2206_Combined.xlsx")

# Calculate Summary Statistics
summary_stats <- GFR_2206 %>%
  group_by(Genotype) %>%
  summarise(
    mean = mean(GFR, na.rm = TRUE),
    sem = sd(GFR, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

# Plot Results
ggplot(GFR_2206, aes(x = Genotype, y = GFR, color = Genotype)) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.7) + # Dot plot
  geom_point(data = summary_stats, aes(y = mean), shape = 18, size = 4, color = "black") + # Mean
  geom_errorbar(data = summary_stats, aes(y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.2, color = "black") + # SEM
  scale_color_manual(values = c("orange", "purple")) +
  ylab("14w GFR (ml/min/100g b.w.)") +
  xlab("Genotype") +
  ggtitle("14w GFR Scores Per Genotype") +
  ylim(0,1500)

# Obtaining P-Value
interaction <- aov(GFR~Genotype, data = GFR_2206)
summary(interaction)
# p-value = 0.676

# Function to calculate R²
get_r2 <- function(model) {
  anova_table <- anova(model)
  ss_total <- sum(anova_table$"Sum Sq")
  ss_model <- sum(anova_table$"Sum Sq"[1])  # Only Genotype term
  r_squared <- ss_model / ss_total
  return(r_squared)
}

# Calculate R²
r2 <- get_r2(interaction)

# Print results
cat("R²:", r2, "\n") # r2 = 0.009383944
```

# Albumin to Creatinine Ratio (ACR) 

ACR at 15 weeks of age.

```{r ACR}
# Albumin to Creatinine Ratio (ACR) @ 15 weeks

# Load Data
ACR_2206_SamePlate_Corrected<-read_excel("~/Library/CloudStorage/OneDrive-TheJacksonLaboratory/Korstanje Lab/22-06 NBL1 KO/22-06 Assays/2206_ACR_SamePlate_Corrected.xlsx")

# Filter Data By Timepoint
ACR_filt <- ACR_2206_SamePlate_Corrected %>%
  filter(Timepoint == "15w") %>%
  mutate(ACR = as.numeric(ACR))

# Calculate Summary Statistics
summary_stats <- ACR_filt %>%
  group_by(Genotype) %>%
  summarise(
    mean = mean(ACR, na.rm = TRUE),
    sem = sd(ACR, na.rm = TRUE) / sqrt(n()),
    .groups = 'drop'
  )

# Plot Results
ggplot(ACR_filt, aes(x = Genotype, y = ACR, color = Genotype)) +
  geom_jitter(width = 0.2, size = 3, alpha = 0.7) + # Dot plot
  geom_point(data = summary_stats, aes(y = mean), shape = 18, size = 4, color = "black") + # Mean
  geom_errorbar(data = summary_stats, aes(y = mean, ymin = mean - sem, ymax = mean + sem), width = 0.2, color = "black") + # SEM
  scale_color_manual(values = c("orange", "purple")) +
  ylab("15w ACR (mg/g)") +
  xlab("Genotype") +
  ggtitle("15w ACR Scores Per Genotype") +
  ylim(0, 12000)

# Obtaining P-Value
interaction <- aov(ACR~Genotype, data = ACR_filt)
summary(interaction)
# p-value = 0.0545

# Function to calculate R²
get_r2 <- function(model) {
  anova_table <- anova(model)
  ss_total <- sum(anova_table$"Sum Sq")
  ss_model <- sum(anova_table$"Sum Sq"[1])  # Only Genotype term
  r_squared <- ss_model / ss_total
  return(r_squared)
}

# Calculate R²
r2 <- get_r2(interaction)

# Print results
cat("R²:", r2, "\n") # r2 = 0.1455244

```