---
title: "NBL1Manuscript_Figure4_A"
author: "Courtney Willey"
date: "2025-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Code Script for Nbl1 Manuscript: NBL1 Correlates with Renal Phenotypes in Mouse Models of Kidney Disease, but is Not Causal.

# Figure 4. RNA Sequencing Analysis of Nbl1 Knockout Mice with Glomerular or Tubular Damage.
XLAS Model

```{r Install}
install.packages("Biobase")
install.packages("limma")
install.packages("edgeR")
install.packages("ggplot2")
install.packages("ggrepel")
install.packages("RColorBrewer")
BiocManager::install("clusterProfiler")
BiocManager::install("org.Mm.eg.db")
BiocManager::install("EnhancedVolcano")
```

```{r Load}
library(Biobase)
library(limma)
library(edgeR)
library(ggplot2)
library(ggrepel)
library(RColorBrewer)
library(clusterProfiler)
library(org.Mm.eg.db)
library(EnhancedVolcano)
```

# Load and Prepare Data
```{r Data}
setwd("~/Library/CloudStorage/OneDrive-TheJacksonLaboratory/Korstanje Lab/22-06 NBL1 KO/22-06 Kidneys/RNAseq/HarshpreetsAnalysis/2206 Combined")
data <-load("Bulk_rnaseq_Nbl1xCol4a5_Nbl1KO_sample_pheno_probeanns.rdata")
ls()

# saving .rdata file as a .txt
#counts <- exprs(cnts.eset)
#tpm <- exprs(tpm.eset)
#fpkm <- exprs(fpkm.eset)

#all(rownames(counts) == rownames(tpm)) # True
#all(rownames(counts) == rownames(fpkm)) # True

#combined <- data.frame(Gene = rownames(counts),
#                       counts, tpm, fpkm)

#colnames(combined)[2:(ncol(counts)+1)] <- paste0("Counts_", colnames(counts))
#colnames(combined)[(ncol(counts)+2):(ncol(counts)+ncol(tpm)+1)] <- paste0("TPM_", #colnames(tpm))
#colnames(combined)[(ncol(counts)+ncol(tpm)+2):ncol(combined)] <- paste0("FPKM_", #colnames(fpkm))


#print(colnames(combined))
#write.table(combined, file = "2206_counts.txt", sep = "\t", quote = FALSE, row.names = FALSE)

probe.anns <- featureData(cnts.eset)@data 
pheno.anns <- phenoData(cnts.eset)@data
pheno.anns$cohort <- factor(pheno.anns$cohort)
pheno.anns$genotype <- factor(pheno.anns$genotype, levels = c("WT", "HET"))

probe.anns.sel <- probe.anns[probe.anns$biotype %in% c(" protein_coding", " lincRNA"),]
cnts <- exprs(cnts.eset)
cnts.sel <- cnts[rownames(probe.anns.sel),] 

dge <- DGEList(cnts.sel)
dge$samples$lib.size <- colSums(dge$counts)
dge$genes <- probe.anns.sel[rownames(dge),]

keep <- filterByExpr(dge, group = pheno.anns$genotype)
dge <- dge[keep,, keep.lib.sizes = FALSE]
dge <- calcNormFactors(dge)
lcpm <- cpm(dge, log = TRUE)  # For visualization only
```

# PCA Before Batch Correction
``` {r PCA}
pca_pre <- prcomp(t(lcpm), scale. = TRUE)
pca_df_pre <- data.frame(PC1 = pca_pre$x[,1], PC2 = pca_pre$x[,2], pheno.anns)

pdf("2206Combined_PCA_BeforeBatch.pdf", width = 10, height = 8)

ggplot(pca_df_pre, aes(x = PC1, y = PC2, color = genotype, shape = cohort)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "PCA Before Batch Correction")

dev.off()
```

# DEG Analysis using limma-voom with Batch Modeled in Design
``` {r DEG}
design <- model.matrix(~ genotype + cohort, data = pheno.anns)
v <- voom(dge, design, plot = FALSE)
fit <- lmFit(v, design)
contrast <- makeContrasts(HETvsWT = genotypeHET, levels = design)
fit2 <- contrasts.fit(fit, contrast)
fit2 <- eBayes(fit2)
```

# PCA After Modeling Batch in Design
``` {r PCA}
exprs_corrected <- removeBatchEffect(v$E, batch = pheno.anns$cohort)
v_corrected <- v
v_corrected$E <- exprs_corrected

pca_post <- prcomp(t(v_corrected$E), scale. = TRUE)
pca_df_post <- data.frame(PC1 = pca_post$x[,1], PC2 = pca_post$x[,2], pheno.anns)

pdf("2206Combined_PCA_AfterBatch.pdf", width = 10, height = 8)

ggplot(pca_df_post, aes(x = PC1, y = PC2, color = genotype, shape = cohort)) +
  geom_point(size = 3) +
  theme_minimal() +
  labs(title = "PCA After Batch Modeled in Design")

dev.off()
```

# Extract DEG Results
``` {r DEG}
deg_results <- topTable(fit2, coef = "HETvsWT", number = Inf, adjust.method = "fdr")
deg_sig <- deg_results[deg_results$P.Value < 0.05, ]
```

# Volcano Plot
``` {r plot}
deg_results$threshold <- "Not Sig"
deg_results$threshold[deg_results$logFC > 0.5 & deg_results$P.Value < 0.1] <- "Up"
deg_results$threshold[deg_results$logFC < -0.5 & deg_results$P.Value < 0.1] <- "Down"

# Label only Htr6 and Nbl1
deg_results$label <- ifelse(deg_results$symbol %in% c("Htr6", "Nbl1", "Ankra2"), deg_results$symbol, "")

# Volcano Plot
pdf("2206Combined_VolcanoPlot_Test_Genes.pdf", width = 10, height = 8)

ggplot(deg_results, aes(x = logFC, y = -log10(P.Value), color = threshold)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("Down" = "red","Not Sig" = "grey", "Up" = "blue")) +
  geom_text_repel(aes(label = label), max.overlaps = 1000) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  coord_cartesian(xlim = c(-3,9), ylim = c(0, 10)) +
  theme_minimal() +
  labs(title = "Volcano Plot: HET vs WT", x = "log2 Fold Change", y = "-log10 P-Value")

dev.off()
```

# Looking at Genes of Interest
``` {r gene}
Htr6 <- "Htr6"
logFC_Htr6 <- deg_results$logFC[deg_results$symbol == Htr6]
Htr6_foldchange<-2^logFC_Htr6
pval_Htr6 <- deg_results$P.Value[deg_results$symbol == Htr6]
#Htr6_foldchange: 13.7141594876157
#logFC_Htr6: 3.7759429999686
#p=8.52475679356065e-05

Nbl1 <- "Nbl1"
logFC_Nbl1 <- deg_results$logFC[deg_results$symbol == Nbl1]
Nbl1_foldchange<-2^logFC_Nbl1
pval_Nbl1 <- deg_results$P.Value[deg_results$symbol == Nbl1]
#Nbl1_foldchange: 0.602923260723498
#logFC_Nbl1: -0.729953705397079
#p=8.42666560711253e-05
```

# Looking at Pathways
``` {r Pathway}
pdf("TopEnrichedPathways.pdf", width = 10, height = 8)

try({
  gene_bitr <- bitr(deg_sig$geneid, fromType = "ENSEMBL", toType = c("ENTREZID", "SYMBOL"), OrgDb = "org.Mm.eg.db")
  keggpath <- enrichKEGG(gene_bitr$ENTREZID, organism = "mmu", pvalueCutoff = 0.05)
  keggpath <- setReadable(keggpath, "org.Mm.eg.db", keyType = "ENTREZID")
  barplot(keggpath, title = "Top Enriched pathways in WT vs HET", font.size = 7)
}, silent = TRUE)

dev.off()

pdf("TopEnrichedBiologicalProcesses.pdf", width = 10, height = 8)

try({
  GOterm.bp <- enrichGO(gene_bitr$ENTREZID, OrgDb = "org.Mm.eg.db", readable = TRUE, ont = "BP")
  dotplot(GOterm.bp, showCategory = 25, font.size = 6, label_format = 50,
          title = "Enriched Biological Processes in WT vs HET")
}, silent = TRUE)

dev.off()
```