---
title: "NBL1Manuscript_Figure6_B"
author: "Courtney Willey"
date: "2025-11-11"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Figure 6. RNA Sequencing Analysis of Nbl1 Knockout Mice with Glomerular or Tubular Damage.
Cisplatin Model

```{r Install}
install.packages("edgeR")
install.packages("sva")
install.packages("limma")
install.packages("ggplot2")
install.packages("Biobase")
install.packages("RColorBrewer")
install.packages("dplyr")
install.packages("ggrepel")
BiocManager::install("PCAtools")
BiocManager::install("EnhancedVolcano")
```

```{r Library}
library(edgeR)
library(sva)
library(limma)
library(ggplot2)
library(Biobase)
library(RColorBrewer)
library(dplyr)
library(ggrepel)
library(PCAtools)
library(EnhancedVolcano)
```

# Load and Prepare Data
``` {r Data}
setwd("~/Library/CloudStorage/OneDrive-TheJacksonLaboratory/Korstanje Lab/22-06 NBL1 KO/22-06 Kidneys/RNAseq/HarshpreetsAnalysis/2105")
load("Bulk_rnaseq_Nbl1KO_cisplatin_sample_pheno_probeanns.rdata")

# saving .rdata file as a .txt
#counts <- exprs(cnts.eset)
#tpm <- exprs(tpm.eset)
#fpkm <- exprs(fpkm.eset)

#all(rownames(counts) == rownames(tpm)) # True
#all(rownames(counts) == rownames(fpkm)) # True

#combined <- data.frame(Gene = rownames(counts),
#                       counts, tpm, fpkm)

#colnames(combined)[2:(ncol(counts)+1)] <- paste0("Counts_", colnames(counts))
#colnames(combined)[(ncol(counts)+2):(ncol(counts)+ncol(tpm)+1)] <- paste0("TPM_", colnames(tpm))
#colnames(combined)[(ncol(counts)+ncol(tpm)+2):ncol(combined)] <- paste0("FPKM_", colnames(fpkm))

#print(colnames(combined))
#write.table(combined, file = "2105_counts.txt", sep = "\t", quote = FALSE, row.names = FALSE)

probe.anns <- featureData(cnts.eset)@data 
pheno.anns <- phenoData(cnts.eset)@data
pheno.anns$treatment <- factor(pheno.anns$treatment, levels = c("PBS", "Cisplatin"))
pheno.anns$genotype <- factor(pheno.anns$genotype, levels = c("WT", "HET"))

probe.anns.sel <- probe.anns[probe.anns$biotype %in% c(" protein_coding", " lincRNA"),]
cnts <- exprs(cnts.eset)
cnts.sel <- cnts[rownames(probe.anns.sel),]

dge <- DGEList(cnts.sel)
dge$samples$lib.size <- colSums(dge$counts)
dge$genes <- probe.anns.sel[rownames(dge),]

keep <- filterByExpr(dge, design = model.matrix(~ genotype * treatment, data = pheno.anns))
dge <- dge[keep,,keep.lib.sizes=FALSE]
dge <- calcNormFactors(dge)
lcpm <- cpm(dge, log = TRUE) #15784,31
```

# Design Matrix and Model Fitting

- The design matrix comprises of the following columns:

- `Intercept` 
- `genotypeHET` 
- `treatmentCisplatin` 
- `genotypeHET.treatmentCisplatin`

`WT` and `PBS` are the reference

The contrast matrix corresponds to the following:

- `HET_Cis_vs_PBS`: treatmentCisplatin = effect of Cisplatin in WT; genotypeHET:treatmentCisplatin = extra effect of Cisplatin in HET compared to WT; So this adds up to the total Cisplatin effect in HET.

- `WT_Cis_vs_PBS`: Is the gene expression in WT Cisplatin different from WT PBS?; Simple — because treatmentCisplatin is the main effect in WT (reference genotype).  

- `HET_vs_WT_PBS`: Is the gene expression in HET PBS different from WT PBS?; This is the main genotype effect in PBS-treated samples (PBS is the reference level for treatment).

- `HET_vs_WT_PBS`: Is the gene expression in HET Cisplatin different from WT Cisplatin?; genotypeHET = baseline HET vs WT under PBS; genotypeHET:treatmentCisplatin = additional difference under Cisplatin; Adds up to total HET vs WT effect under Cisplatin.  
Interaction    = genotypeHET.treatmentCisplatin

- `Interaction`: Does the effect of Cisplatin differ between HET and WT?; This is the statistical interaction term. If this is significant, it means: The treatment response is not consistent across genotypes.

```{r design}
design <- model.matrix(~ genotype * treatment, data = pheno.anns)
colnames(design) <- make.names(colnames(design))

dge <- estimateDisp(dge, design)
fit <- glmQLFit(dge, design)

contrast.matrix <- makeContrasts(
  # Effect of Cisplatin vs PBS in HET
  HET_Cis_vs_PBS = treatmentCisplatin + genotypeHET.treatmentCisplatin,
  # Effect of Cisplatin vs PBS in WT
  WT_Cis_vs_PBS  = treatmentCisplatin,
  # HET vs WT under PBS
  HET_vs_WT_PBS  = genotypeHET,
  # HET vs WT under Cisplatin
  HET_vs_WT_Cis  = genotypeHET + genotypeHET.treatmentCisplatin,
  # Interaction (does treatment effect differ by genotype)
  Interaction    = genotypeHET.treatmentCisplatin,
  levels = design
)
```

# Differential Expression Results

```{r DEG}
res_HET_Cis_vs_PBS <- glmQLFTest(fit, contrast = contrast.matrix[, "HET_Cis_vs_PBS"])
res_WT_Cis_vs_PBS <- glmQLFTest(fit, contrast = contrast.matrix[, "WT_Cis_vs_PBS"])
res_HET_vs_WT_PBS <- glmQLFTest(fit, contrast = contrast.matrix[, "HET_vs_WT_PBS"])
res_HET_vs_WT_Cis <- glmQLFTest(fit, contrast = contrast.matrix[, "HET_vs_WT_Cis"])

toptags_HET_Cis_vs_PBS <- topTags(res_HET_Cis_vs_PBS, n = Inf)$table %>% filter(FDR < 0.05) #10318 sig genes; Htr6 upregulated
toptags_WT_Cis_vs_PBS <- topTags(res_WT_Cis_vs_PBS, n = Inf)$table %>% filter(FDR < 0.05) #10826 sig genes
toptags_HET_vs_WT_PBS <- topTags(res_HET_vs_WT_PBS, n = Inf)$table %>% filter(FDR < 0.05) #0
toptags_HET_vs_WT_Cis <- topTags(res_HET_vs_WT_Cis, n = Inf)$table %>% filter(FDR < 0.05) #1 gene Htr6
toptags_HET_vs_WT_Cis_P <- topTags(res_HET_vs_WT_Cis, n = Inf)$table %>% filter(PValue < 0.05) #1056 sig
```

# Library Size Plot
```{r Plot}
librarySizes <- colSums(dge$counts)
libData <- data.frame(
  Sample = names(librarySizes),
  LibrarySize = librarySizes,
  Unique_id = pheno.anns$Unique_id,
  Genotype = pheno.anns$genotype
)

ggplot(libData, aes(x = Unique_id, y = LibrarySize, fill = Genotype)) +
  geom_bar(stat = "identity", position = "dodge") +
  xlab("Sample Unique ID") +
  ylab("Library Size") +
  labs(fill = "Genotype") +
  ggtitle("Library Sizes for Bulk RNAseq Kidney samples") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

# PCA Plot
```{r PCA}
pcp <- pca(lcpm, metadata = pheno.anns, removeVar = 0.1)

pdf("2105_PCA.pdf", width = 10, height = 8)

biplot(pcp,
       showLoadings = FALSE,
       legendPosition = 'right',
       labSize = 3,
       pointSize = 5,
       sizeLoadingsNames = 3,
       max.overlaps = Inf,
       colby = 'genotype',
       shape = 'treatment',
       lab =NULL)

dev.off()
```

# Htr6 Expression Boxplot
```{r Htr6}
# Extract Htr6 expression
htr6_expr <- lcpm["ENSMUSG00000028747_Htr6", , drop = FALSE]

# Build plotting dataframe
plot_df <- cbind(
  expression = as.numeric(htr6_expr),
  pheno.anns[, c("genotype", "treatment")]
)
plot_df$group <- interaction(plot_df$genotype, plot_df$treatment, sep = "_")

pdf("2105_Htr6ExpressionbyGroup.pdf", width = 10, height = 8)

ggplot(plot_df, aes(x = group, y = expression, fill = group)) +
  geom_boxplot() +
  geom_jitter(width = 0.2, size = 2) +
  labs(
    title = "Htr6 expression across groups",
    x = "Group",
    y = "log2 CPM"
  ) +
  scale_fill_brewer(palette = "Set2") +
  theme_minimal()

dev.off()
```

# Volcano Plot
```{r Volcano}
label_genes <- toptags_HET_Cis_vs_PBS$symbol[
  toptags_HET_Cis_vs_PBS$logFC >= 3 | toptags_HET_Cis_vs_PBS$symbol == "Htr6"]

#Ensure label list has no NAs or duplicates
label_genes <- unique(na.omit(label_genes))

pdf("2105_VolcanoPlot.pdf", width = 10, height = 8)

EnhancedVolcano(
  toptags_HET_Cis_vs_PBS,
  lab = toptags_HET_Cis_vs_PBS$symbol,
  selectLab = label_genes,
  x = "logFC",
  y = "FDR",
  pCutoff = 0.05,
  FCcutoff = 3.0,
  pointSize = 3.0,
  labSize = 5.0,
  drawConnectors = TRUE,
  widthConnectors = 0.75,
  colAlpha = 0.6,  # slightly transparent points
  legendLabels = c("Not Significant", "log2FC", "FDR", "FDR & log2FC"),
  ylab = bquote(-Log[10]~FDR),
  title = "Genes with log2FC ≥ 3 + Htr6"
)

dev.off()
```

# Looking at Cisplatin WT vs Cisplatin HET
``` {r cisplatin}
# Load full results
toptags_HET_vs_WT_Cis_P <- topTags(res_HET_vs_WT_Cis, n = Inf)$table

# Define thresholds
toptags_HET_vs_WT_Cis_P$threshold <- "Not Sig"
toptags_HET_vs_WT_Cis_P$threshold[toptags_HET_vs_WT_Cis_P$logFC > 0.5 & toptags_HET_vs_WT_Cis_P$PValue < 0.05] <- "Up"
toptags_HET_vs_WT_Cis_P$threshold[toptags_HET_vs_WT_Cis_P$logFC < -0.5 & toptags_HET_vs_WT_Cis_P$PValue < 0.05] <- "Down"

# Label top 10 "Not Sig" genes by lowest FDR
not_sig_genes <- toptags_HET_vs_WT_Cis_P[toptags_HET_vs_WT_Cis_P$threshold == "Not Sig", ]
top10_not_sig <- head(not_sig_genes[order(not_sig_genes$PValue), ], 10)

# Label only Htr6 and Nbl1
toptags_HET_vs_WT_Cis_P$label <- ifelse(
  toptags_HET_vs_WT_Cis_P$symbol %in% c("Htr6", "Nbl1", "Cnih2", "Kng1"),
  toptags_HET_vs_WT_Cis_P$symbol,
  ""
)

# Ensure threshold is a factor with correct order
toptags_HET_vs_WT_Cis_P$threshold <- factor(
  toptags_HET_vs_WT_Cis_P$threshold,
  levels = c("Down", "Not Sig", "Up")
)

# Plot
pdf("2105_VolcanoPlot_HET_vs_WT_Cis_ggplot_Pvalue_Test.pdf", width = 10, height = 8)

ggplot(toptags_HET_vs_WT_Cis_P, aes(x = logFC, y = -log10(PValue), color = threshold)) +
  geom_point(alpha = 0.7) +
  scale_color_manual(values = c("Down" = "red", "Not Sig" = "grey", "Up" = "blue")) +
  geom_text_repel(aes(label = label), max.overlaps = 1000) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "black") +
  coord_cartesian(xlim = c(-3,9), ylim = c(0, 10)) +
  theme_minimal() +
  labs(
    title = "Volcano Plot: HET vs WT Treated with Cisplatin",
    x = "log2 Fold Change",
    y = "-log10 P-Value"
  )

dev.off()
```

# Looking at Genes of Interest
``` {r Genes}
Htr6 <- "Htr6"
logFC_Htr6 <- toptags_HET_vs_WT_Cis_P$logFC[toptags_HET_vs_WT_Cis_P$symbol == Htr6]
Htr6_foldchange<-2^logFC_Htr6
pval_Htr6 <- toptags_HET_vs_WT_Cis_P$PValue[toptags_HET_vs_WT_Cis_P$symbol == Htr6]
#Htr6_foldchange: 130.648445926209
#logFC_Htr6: 7.02954615356097
#p=2.11094739860647e-10

Nbl1 <- "Nbl1"
logFC_Nbl1 <- toptags_HET_vs_WT_Cis_P$logFC[toptags_HET_vs_WT_Cis_P$symbol == Nbl1]
Nbl1_foldchange<-2^logFC_Nbl1
pval_Nbl1 <- toptags_HET_vs_WT_Cis_P$PValue[toptags_HET_vs_WT_Cis_P$symbol == Nbl1]
#Nbl1_foldchange: 0.354343958332172
#logFC_Nbl1: -1.49677764401211
#p=0.000263732312594367
```
```